---
title: "Analytics Edge - Rebalancing Bluebikes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Set path to your work directory
library(plyr)
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(magrittr)
library(ggplot2)
library(caret)
library(glmnet)
library(lars)
library(leaps)
library(gbm)
library(tibble)
library(ROCR)
library(shiny)
```

# Clasic datasets

```{r}
rides_201909 = read.csv("201909-bluebikes-tripdata.csv")
# rides_201904 = read.csv("201904-bluebikes-tripdata.csv")
# rides_201905 = read.csv("201905-bluebikes-tripdata.csv")
# rides_201906 = read.csv("201906-bluebikes-tripdata.csv")
# rides_201907 = read.csv("201907-bluebikes-tripdata.csv")
# rides_201908 = read.csv("201908-bluebikes-tripdata.csv")

df <- rides_201909 %>% mutate(starttime = as.POSIXct(starttime),
                              stoptime = as.POSIXct(stoptime),
                              gender = as.factor(gender),
                              start.year = as.integer(format(starttime, "%Y")),
                              start.month = as.integer(format(starttime, "%m")),
                              start.day.of.month = as.integer(format(starttime, "%d")),
                              start.day.of.week = as.integer(format(starttime, "%w")),
                              start.hour = as.integer(format(starttime, "%H")),
                              start.min = as.integer(format(starttime, "%M")),
                              start.time.of.day = start.hour + (start.min/60), 
                              start.quarter = as.integer(start.time.of.day*4)/4,
                              .keep = "unused") %>%
  mutate(
    start.time.interval = floor_date(starttime, "15 min"),
    .after = "starttime"
  ) %>%
  mutate(
    stop.time.interval = floor_date(stoptime, "15 min"),
    .after= 'stoptime'
    )

# Merging df with stations to get number of docks

stations = read.csv("current_bluebikes_stations.csv")
colnames(stations) = stations[1,]
stations = stations[2:dim(stations)[1],] %>% rename(docks = 'Total docks')
glimpse(stations)
df <- df %>%
    merge(stations %>% select(Name, docks),
        by.x = 'end.station.name',
        by.y = 'Name',
        suffixes = c('','.end'), all.x = TRUE) %>%
  merge(stations %>% select(Name, docks),
        by.x = 'start.station.name',
        by.y = 'Name',
        suffixes = c('','.start'), all.x = TRUE) %>%
  mutate(docks.start= as.numeric(docks.start),
         docks.end = as.numeric(docks),
         .keep = 'unused'
         )

# Getting the station.id from the df dataframe per station - dont know why it is NOT in the station file.

stations <- stations %>%
  merge(df %>% select(station.name = start.station.name, station.id = start.station.id) %>%
          group_by(station.name, station.id) %>% summarise(total.rides = n()),
        by.x = 'Name',
        by.y = 'station.name',
        all.x = TRUE) %>%
  arrange(desc(total.rides))
```

```{r}
quantile(df$tripduration, c(0.25,0.5,0.75, .9))
```


## Top 10 stations

```{r}
stations %>% head(10)
```

## Visualizing number of rides
```{r}
df_count_rides <- df %>% filter(start.day.of.month <= 28) %>%
  group_by(start.quarter, start.day.of.week) %>%
  summarise(number.of.rides = n()) %>%
  ungroup()

summary(df_count_rides)
```

```{r}
# Scatter plot by day of week
ggplot(data=df_count_rides,aes(x=start.quarter)) +
  geom_point(aes(y=number.of.rides, color=factor(start.day.of.week)),size=.5) +
  theme_bw() +
  ylab("Number of rides across the city") +
  theme(axis.title=element_text(size=10), axis.text=element_text(size=10), legend.title=element_text(size=10), legend.text=element_text(size=12)) + 
  scale_color_brewer(name='Weekday',labels=c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday'), palette="Set1")
```

## Computing departure, arrival and net arrival rate per station per time interval

```{r}
# Compute the net activity per hour 

df.departure <- df %>% group_by(start.time.interval, start.station.name) %>%
  summarise(number.of.departures = n()) %>%
  ungroup()

df.arrival <- df %>% group_by(stop.time.interval, end.station.name) %>%
  summarise(number_of_arrival = n()) %>%
  ungroup()

# This df contains net arrival rate per station per time interval.
df.net.arrival <- df.departure %>% 
  merge(df.arrival,
        by.x = c('start.station.name','start.time.interval'),
        by.y = c('end.station.name','stop.time.interval'),
        all.x = TRUE,
        all.y = TRUE
  ) %>%
  rename(station.name = start.station.name)

df.net.arrival <- df.net.arrival %>%
  complete(start.time.interval = seq(as.POSIXct('2019-09-01 00:00:00:0000'), 
                                     as.POSIXct('2019-10-01 00:00:00:0000'), by = '15 min'), station.name) %>%
  merge(stations %>% select(-c(Number, total.rides)),
        by.x = 'station.name',
        by.y = 'Name',
        all.x = TRUE) %>%
  mutate(number_of_arrival = replace_na(number_of_arrival,0),
         number.of.departures = replace_na(number.of.departures,0),
         net.arrival.rate = number_of_arrival - number.of.departures) %>%
  arrange(station.name, start.time.interval)
```

```{r}
### Heuristic for empty/full stations
# Play on parameters to have something more smooth!!!

# Huge problem: Some time slots are both empty and nonempty.

## For empty stations
df.net.arrival <- df.net.arrival %>% 
  group_by(station.name) %>%
  mutate(empty = 0) %>%
  mutate(empty = case_when((lag(net.arrival.rate) <= -2) & (net.arrival.rate >= -1) & (net.arrival.rate <= 2) ~ 1,
                           TRUE ~ 0)) %>%
  ungroup()

# Careful - this is long - I don't know how to make it faster
# What I do is that I propagate the empty only one by one - trying a loop was way slower
# 96 is for a day (1 day = 24 * 4 quarters = 96 quarters)
for (i in 1:96){
  df.net.arrival <- df.net.arrival %>%
    group_by(station.name) %>%
    mutate(empty = case_when((lag(empty) == 1) & (net.arrival.rate <= 1) & (net.arrival.rate >= -1) ~ 1,
                             empty == 1 ~ 1,
                             TRUE ~ 0))
}

## For full stations
df.net.arrival <- df.net.arrival %>% 
  group_by(station.name) %>%
  mutate(full = 0) %>%
  mutate(full = case_when((lag(net.arrival.rate) >= 2) & (net.arrival.rate <= 1) & (net.arrival.rate >= -2) ~ 1,
                           TRUE ~ 0)) %>%
  ungroup()

# Careful - this is long - I don't know how to make it faster
for (i in 1:96){
  df.net.arrival <- df.net.arrival %>%
    group_by(station.name) %>%
    mutate(full = case_when((lag(full) == 1) & (net.arrival.rate >= -1) & (net.arrival.rate <= 1) ~ 1,
                             full == 1 ~ 1,
                             TRUE ~ 0))
}
```

```{r}
view(df.net.arrival %>% filter(station.name == 'Kendall T'))
```











## Computing the net evolution of number of bikes in a station during the day
```{r}
# Approach: Build a database for each station, where events per this station are listed.

start.event.stations = df %>%
  select(station.id = start.station.id, station.name = start.station.name, time = starttime, 
         time.interval = start.time.interval, year = start.year, month = start.month, 
         day.of.month = start.day.of.month, day.of.week = start.day.of.week, hour = start.hour, 
         min = start.min, tripduration, station.latitude = start.station.latitude,
         station.longitude = start.station.longitude, bikeid, usertype, docks = docks.start) %>%
  mutate(start.ind = 1, end.ind = 0)

end.event.stations = df %>% mutate(year = as.integer(format(stoptime, "%Y")),
                                   month = as.integer(format(stoptime, "%m")),
                                   day.of.month = as.integer(format(stoptime, "%d")),
                                   day.of.week = as.integer(format(stoptime, "%w")),
                                   hour = as.integer(format(stoptime, "%H")),
                                   min = as.integer(format(stoptime, "%M"))) %>%
  select(station.id = end.station.id, station.name = end.station.name, time = stoptime, 
         time.interval = stop.time.interval, year, month, day.of.month, day.of.week, hour, min,
         tripduration, station.latitude = end.station.latitude,
         station.longitude = end.station.longitude, bikeid, usertype, docks = docks.end) %>%
  mutate(end.ind = 1, start.ind = 0)

event.stations = start.event.stations %>%
  merge(end.event.stations, all.x = TRUE, all.y = TRUE) %>%
  arrange(time) %>%
  group_by(day.of.month, station.id) %>%
  mutate(cum.departures = cumsum(replace_na(start.ind, 0)),
         cum.arrivals = cumsum(replace_na(end.ind, 0)),
         net.daily.arrivals = cum.arrivals - cum.departures)
```


```{r}
# Scatter plot by day
ggplot(data=event.stations %>% filter(station.name == 'Kendall T')) +
  geom_line(aes(x =time, y=net.daily.arrivals, color = factor(day.of.month))) +
  theme_bw() +
  ylab("Cumulative net arrivals") +
  theme(axis.title=element_text(size=10), axis.text=element_text(size=10), legend.title=element_text(size=10), legend.text=element_text(size=12))
```

```{r}
# Scatter plot by day - Change the name of the station
name = "Dana Park"

ggplot(data=event.stations %>% filter(station.name == name) %>% 
         mutate(week.time = min+60*hour+1440*day.of.week)) +
  geom_line(aes(x = week.time, y=net.daily.arrivals, color = factor(day.of.month))) +
  geom_hline(aes(yintercept=0*as.numeric((stations %>% filter(Name == name))$docks)), color = "red") +
  geom_hline(aes(yintercept=-as.numeric((stations %>% filter(Name == name))$docks)), color = "red") +
  theme_bw() +
  ylab("Cumulative net arrivals") +
  theme(axis.title=element_text(size=10), axis.text=element_text(size=10), legend.title=element_text(size=10), legend.text=element_text(size=12))
```

```{r}
# station.names = unique(event.stations$station.name)
# station.names = sample(station.names,5)
station.names = c("Kendall T",
                  "One Broadway / Kendall Sq at Main St / 3rd St",
                  "Third at Binney",
                  "Kendall Street",
                  "One Memorial Drive")

for (name in station.names) {
  plt = ggplot(data=event.stations %>% filter(station.name == name) %>%
                 mutate(week.time = min+60*hour+1440*day.of.week)) +
    geom_line(aes(x = week.time, y=net.daily.arrivals, color = factor(day.of.month))) +
    geom_hline(aes(yintercept=as.numeric((stations %>% filter(Name == name))$docks)), color = "red") +
    geom_hline(aes(yintercept=-as.numeric((stations %>% filter(Name == name))$docks)), color = "red") +
    theme_bw() +
    ylab("Cumulative net arrivals") +
    labs(title = name) +
    theme(axis.title=element_text(size=10), axis.text=element_text(size=10), 
          legend.title=element_text(size=10), legend.text=element_text(size=12))
  print(plt)
}
```

## Network flows through the day

```{r}
### Current time.interval of 1h
network_flows <- rides_201909 %>% mutate(starttime = as.POSIXct(starttime),
                              stoptime = as.POSIXct(stoptime),
                              gender = as.factor(gender),
                              start.year = as.integer(format(starttime, "%Y")),
                              start.month = as.integer(format(starttime, "%m")),
                              start.day.of.month = as.integer(format(starttime, "%d")),
                              start.day.of.week = as.integer(format(starttime, "%w")),
                              start.hour = as.integer(format(starttime, "%H")),
                              start.min = as.integer(format(starttime, "%M")),
                              start.time.of.day = start.hour + (start.min/60), 
                              start.quarter = as.integer(start.time.of.day*4)/4,
                              .keep = "unused") %>%
  mutate(                                                                     # Intervals of 1h
    start.time.interval = floor_date(starttime, "1h"),
    .after = "starttime"
  ) %>%
  mutate(
    stop.time.interval = floor_date(stoptime, "1h"),
    .after= 'stoptime'
    ) %>%
  group_by(start.station.name, end.station.name, start.time.interval) %>%
  summarise(hourly.dep.flow = n()) %>%                                        # Now adding departure_flow
  merge(stations %>% select(Name, docks),                                     # Adding number of docks per station
        by.x = 'start.station.name',
        by.y = 'Name',
        suffixes = c('','.start'), all.x = TRUE) %>%
  merge(stations %>% select(Name, docks),
        by.x = 'end.station.name',
        by.y = 'Name',
        suffixes = c('','.end'), all.x = TRUE) %>%
  rename(start.docks = docks, end.docks = docks.end) %>%
  arrange(start.station.name, start.time.interval, end.station.name) %>%
  group_by(start.station.name, start.time.interval) %>%                       # Now adding total_flows
  mutate(hourly.dep.flow.total = sum(hourly.dep.flow)) %>%
  ungroup() %>%
  mutate(hourly.dep.flow.pct = hourly.dep.flow / hourly.dep.flow.total)
  ## If needed, add missing time interval and stations and populate with 0 network flows.

view(network_flows)
```

Understand the life of one bike for a day.
Let's say I start from Central.
At each hour, I look

### Other idea: bottom-up approach

```{r}
# There are 'only' 3496 bikes for 300ish stations - 10 bikes per station
bikes = unique(df$bikeid)
length(unique(df$bikeid))
```

Look at each bike id throughout the day, where it started and to where is his last trip. Thera are 'only' 3496 different bikes in our dataset.

I want, for each bike in the dataset:

* the station it departed from first time and time associated

* I don't exactly care about the stations it goes through, however I could try to look at the number of stations

* the station it arrived in the evening and time associated

```{r}
# First let's look at May 1st, 2019
first.day <- df %>% filter(start.day.of.month == 1) %>% arrange(starttime)

for (bike in bikes){
  for (row in 1:nrow(first.day)){
    # date  <- stock[row, "date"]
    
  }
}

```






















































# Old Stuff

## Releasability and dockability

```{r}
df = read.csv("Rideability.csv")

# Appropriate names
for (i in 3:ncol(df)){
  names(df)[i] <- paste(names(df)[i], sub("Unweighted ", "", df[1,i]), sep = ".")
  }
df = df[2:nrow(df),2:ncol(df)]

# Change data type
percent_to_num <- function(x) as.numeric(sub("%", "", x))
df <- df %>% mutate(
  time.interval = as.POSIXct(Name), 
  .keep = "unused", 
  .before = "Beacon.St.at.Massachusetts.Ave.Dockability"
  ) %>%
  mutate_if(is.character, percent_to_num) %>%
  mutate(year = as.integer(format(time.interval, "%Y")),
         month = as.integer(format(time.interval, "%m")),
         day.of.month = as.integer(format(time.interval, "%d")),
         day.of.week = as.integer(format(time.interval, "%w")),
         hour = as.integer(format(time.interval, "%H")),
         min = as.integer(format(time.interval, "%M")),
         time.of.day = hour + (min/60),
         .before = "Beacon.St.at.Massachusetts.Ave.Dockability")

train <- df %>% filter(month == 6, year == 2019)

summary(train)
```

```{r}
# Scatter plot by day of week
ggplot(data=train ,aes(x=time.of.day)) +
  geom_point(aes(y=Beacon.St.at.Massachusetts.Ave.Dockability,color=factor(day.of.week)),size=.5) +
  theme_bw() +
  xlab('Time of day') +
  ylab("Dockability at Beacon.St.at.Massachusetts.Ave") +
  theme(axis.title=element_text(size=10), axis.text=element_text(size=10), legend.title=element_text(size=10), legend.text=element_text(size=12)) + 
  scale_color_brewer(name='Weekday',labels=c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday'), palette="Set1")
```

```{r}
# Scatter plot by day of week
ggplot(data=train %>% filter(day.of.week == 5),aes(x=time.of.day)) +
  geom_point(aes(y=MIT.at.Mass.Ave...Amherst.St.1.Releasability,color=factor(day.of.week)),size=.5) +
  theme_bw() +
  xlab('Time of day') +
  ylab("Releasability at Mass.Ave...Amherst.St.1.") +
  theme(axis.title=element_text(size=10), axis.text=element_text(size=10), legend.title=element_text(size=10), legend.text=element_text(size=12)) + 
  scale_color_brewer(name='Weekday',labels=c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday'), palette="Set1")
```

```{r}
# This next cell is old stuff

# See what bikes coming in and out a station look like
# hv.station.start <- df %>% filter(start.station.id == 149) %>%
#   mutate(ind = 1) %>%
#   arrange(starttime) %>%
#   group_by(start.day.of.month) %>%
#   mutate(cum.arrivals = cumsum(ind))
# 
# hv.station.end <- df %>% filter(end.station.id == 149) %>%
#   mutate(ind = 1) %>%
#   arrange(starttime) %>%
#   group_by(start.day.of.month) %>%
#   mutate(cum.departures = cumsum(ind))
# 
# hv.station <- df %>% filter(start.station.id == 149) %>%
#   mutate(start.ind = 1) %>%
#   merge(df %>% filter(end.station.id == 149) %>%
#           mutate(end.ind = 1),
#         all.x = TRUE, all.y = TRUE) %>%
#   group_by(start.day.of.month) %>%
#   arrange(starttime) %>%
#   mutate(cum.departure = cumsum(replace_na(start.ind, 0)),
#          cum.arrival = cumsum(replace_na(end.ind,0))) %>%
#   mutate(net.cum.arrivals = cum.arrival - cum.departure)
```



